Analyzing DECREASE trials to estimate evidence of data manipulation

### 2017-07-27 submission/manuscript.tex
Revision round gazilion

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Analyzing DECREASE trials to estimate evidence of data manipulation},
            pdfauthor={Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper, Markus Hollman},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials to estimate evidence of data manipulation}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper, Markus Hollman}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{14 July, 2017}

@@

@@

The trials subject to concerns regarding scientific integrity were the
Dutch DECREASE-I and DECREASE-IV trials\textsuperscript{4--6}. The
committees that investigated the integrity of the DECREASE trials
reported that data manipulation was likely but that the extent of the
data manipulation remained unclear\textsuperscript{4--6}. Moreover, the
latest guidelines still recommend the usage of beta-blockers in the
perioperative period in certain cases\textsuperscript{10,11}, where some
of these guidelines are based on other work by the PI of the DECREASE
trials. Considering the potential harmful consequences of guidelines
based on work by someone who has been repeatedly investigated for
breaching scientific integrity, we aim to estimate the extent of data
manipulation in the DECREASE studies\textsuperscript{2--6} to further
stimulate the debate on using beta-blockers in the perioperative period
patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data but did not investigate the extent to
which the DECREASE trials deviated from comparable trials. Provenance is
primarily concerned with the origins of the data, verifying things such
as (but not limited to) the informed consent and whether data
corresponded to patient files. The committee reports did not neglect
statistical evaluation however: according to the report a statistical
expert evaluated the applicability of forensic statistical
methods\textsuperscript{6} to evaluate results of trials separately
(i.e., DECREASE-I, DECREASE-IV) although it lacks details as to how this
evaluation took place. The expert concluded that the previously applied
methods by him were not applicable for use in this case. Nonetheless, in
this report we compare across trials, which is a method that has
previously been used to monitor trial data quality or to test for
potential data anomalies\textsuperscript{12,13}. Moreover, comparing
across trials instead of evaluating them separately has previously
proven to be effective in detecting data
manipulation\textsuperscript{14}. Comparing the DECREASE trials to other
published trials studying the effectiveness of beta-blockers with
respect to perioperative mortality could prove informative of the
potential extent of the manipulation in the DECREASE trials.

@@

To statistically investigate the evidence for data manipulation in the
DECREASE studies\textsuperscript{2,3}, we took three steps. First, we
reproduced the findings from the 2014 meta-analysis by Bouri et
al.\textsuperscript{9}, which contained sufficient information to
estimate the deviation of the DECREASE trials from other published
trials on beta-blockers. We also included type of beta-blocker to
inspect whether this is predictive of the effect of beta-blockers on
perioperative mortality. Second, we evaluated the probability that the
DECREASE trials occurred assuming no data manipulation. Third, we
reversed this assumption and assumed that data manipulation did occur
and estimated how many data points would have to be manipulated in order
to reproduce the results of the DECREASE trials, if the other published
trials are regarded as estimating the true effect of beta-blockers on
perioperative mortality in patients undergoing non-cardiac surgery.
Considering the committees investigating the scientific integrity of the
DECREASE trials was unable to assess this, we consider it worthwhile to
investigate this further.

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled
these using the \texttt{R} package \texttt{metafor}\textsuperscript{16}.
We estimated a weighted random-effects model using the restricted
maximum-likelihood estimator (i.e., \texttt{REML})\textsuperscript{17}
to estimate the variance of effects. We used the default weighting
procedure in the \texttt{metafor} package. We added 0.5 to each cell
count, as is common in meta-analyses on risk- and odds ratios in order
to prevent computational artefacts {[}18; p397{]}. The 2014
meta-analysis\textsuperscript{9} did not specify the variance estimate
used; hence, minor discrepancies between our estimates and the original
estimates could be due to differences in the estimation procedure.

@@

We were able to closely reproduce the estimates for the different sets
of studies (Figure 2 of the 2014 meta-analysis\textsuperscript{9}).
Bouri et al. differentiated between the estimates from the non-DECREASE
trials (\(k=9\)) and the DECREASE trials (\(k=2\)). We confirmed the
effect size estimates and the variance estimates for both the
non-DECREASE- and the DECREASE trials, except for some minor
discrepancies at the second decimal level. Table 1 depicts the original
and reproduced values for both sets of studies.

\begin{longtable}[]{@{}lllll@{}}
\caption{The original- and reproduced meta-analytic results based on the
data provided in the 2014 meta-analysis by Bouri et al.}\tabularnewline
\toprule
& & Risk ratio & \(\tau^2\) & Confidence interval\tabularnewline
Non-DECREASE (k=9) & Original & 1.27 & 0 & {[}1.01;
1.60{]}\tabularnewline
& Reproduced & 1.27 & 0 & {[}1.01; 1.61{]}\tabularnewline
DECREASE (k=2) & Original & 0.42 & 0.29 & {[}0.15;
1.23{]}\tabularnewline
& Reproduced & 0.47 & 0.14 & {[}0.19; 1.15{]}\tabularnewline
\bottomrule
\end{longtable}

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to reproduce results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{9}.
Surprisingly, our results showed a bit more evidence against equal
subgroups than the original meta-analysis\textsuperscript{9} (original:
\(\chi^2(1)=3.91,p=.05\); reproduced: \(\chi^2(1)=6.12,p=0.013\)).
Additionally, the original analyses showed substantial residual
heterogeneity (\(I^2=74.4\)\%) whereas we found no residual
heterogeneity (\(I^2=0\)\%). Different variance estimates (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors (including a reminder after several weeks) but did
not receive a response. Nonetheless, the broad strokes of the
meta-regression confirmed that the DECREASE trials were the determining
predictor for the effectiveness of beta-blockers (including DECREASE:
\(RR=0.509\); excluding DECREASE: \(RR=1.275\)).

Additionally, and exploratively, we evaluated the predictive effect of
the type of beta-blocker used in the trials. The DECREASE trials
remained predictive of decreased mortality (\(RR=0.509\)), whereas the
non-DECREASE trials provide tentative, but uncertain, evidence that
atenolol results in lower mortality (\(RR=0.777\)). Nonetheless, for
other beta-blockers in the non-DECREASE trials, there is still tentative
and uncertain evidence that beta-blockers increase mortality
(bisoprolol: \(RR=2.973\); metoprolol: \(RR=1.303\); propranolol:
\(RR=1.7\)). Table 2 shows the meta-regression results in full.

\begin{longtable}[]{@{}lll@{}}
\caption{Meta-regression results (k=11) for log(RR), including dummy
predictors for DECREASE trials (reference: non-DECREASE trials) and type
of beta-blockers used in the trial (reference:
atenolol).}\tabularnewline
\toprule
& Estimate & 95\% CI\tabularnewline
Intercept & -0.252 & -1.228; 0.724\tabularnewline
Non-DECREASE & - & -\tabularnewline
DECREASE & -1.765 & -5.05; 1.519\tabularnewline
Atenolol & - & -\tabularnewline
Bisoprolol & 1.342 & -2.015; 4.698\tabularnewline
Metoprolol & 0.517 & -0.49; 1.523\tabularnewline
Propranolol & 0.783 & -1.5; 3.065\tabularnewline
\bottomrule
\end{longtable}

@@

Based on the non-DECREASE estimates from Step 1, we estimated the
probability of obtaining the results in the DECREASE trials. To this
end, we assumed that the non-DECREASE trials provide a valid
representation of the true effect of beta-blockers on perioperative
mortality (similar to Bouri et al.\textsuperscript{9}). The estimated
probability of the observed data under a given true effect is also known
as the veracity of the data\textsuperscript{19}. We assumed that the
non-DECREASE studies estimated the true effect distribution of
perioperative beta-blockade on mortality, not perturbed by publication
bias due to statistical (non)significance. Publication bias was assumed
to not be a problem because a substantial number of nonsignificant
effects are included in the dataset (9 of 11 results are
nonsignificant).

@@

Based on the estimated mean log RR and its credible interval in the
non-DECREASE studies, we computed the probability of the observed log RR
in the DECREASE trials. The estimates of the non-DECREASE studies were
obtained from Step 1, which include the estimated log RR (i.e., 0.24),
and its 95\% credibility interval as provided by the package
\texttt{metafor} (i.e., {[}0.008; 0.477{]}). The meta-analysis model
assumes a normal distribution of population effects with the estimated
effect as the mean of the distribution. The 95\% credibility interval
denotes the bounds of the normal distribution that covers 95\% of the
density, where the standard deviation is calculated as the distance from
the mean to either bound, divided by 1.96. This allows for an
approximation of the population effect distribution, as depicted in
Figure 1.

{\centering \includegraphics[width=0.8\linewidth]{../figures/fig1}

\caption{Density plot of the estimated true effect distribution based on the non-DECREASE studies only, with the position of the DECREASE studies highlighted.}\label{fig:figure 1}
\end{figure}

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the \(p\)-value for the null
hypothesis that the DECREASE trials arise from the same effect
distribution as the non-DECREASE trials. This assumes that the
information available from the other trials is informative of the true
population effect.

@@

Figure 2 indicates that the DECREASE trials are highly unlikely under
the estimated effect distribution based on the non-DECREASE trials. More
specifically, the results from DECREASE-I (or more extreme) have a
probability of \(6.9621238\times 10^{-45}\) (approximately 7 in a
\href{https://en.wikipedia.org/wiki/Names_of_large_numbers}{quattuordecillion};
\(z=-14.057\)) and the results from DECREASE-IV have a probability of
\(6.556057\times 10^{-9}\) (approximately 7 in a billion; \(z=-5.802\)).
This indicates that the DECREASE trial results are unlikely to have come
from the same population effect distribution as the non-DECREASE trials.
Moreover, observing two of such extremely unlikely results jointly, as
in the DECREASE trials, is nearly impossible,
\(4.564408\times 10^{-53}\). Hence, this result indicates that the
DECREASE trials are severely different from the non-DECREASE trials.

Results from Step 1 indicated that no between-trial variance (i.e.,
homogeneity; \(\tau^2=0\)) of the effects was observed; given the small
number of trials included (i.e., 9) this estimate is highly uncertain,
however. The total \(N\) across the non-DECREASE trials was 10529.We
conducted sensitivity analyses to see how dependent results are on the
heterogeneity estimate (not preregistered;
\href{https://osf.io/vnmzc}{osf.io/vnmzc}). The probability of observing
the DECREASE trials stays approximately below 1 out of 1000 until the
variance estimate is 0.25; the probability stays approximately below 1
out of 100 until the variance estimate is 0.43 (see Figure 2). To put
these numbers into context, a variance of 0.25 would suggest that
results of perioperative beta-blockade vary substantially due to
contextual circumstances of the study, even if perioperative
beta-blockade has no effect whatsoever (RRs between 0.779 and 1.284 in
\textasciitilde{}64\% of the cases).

@@

We estimated the number of data points that would need to be manipulated
to arrive at the estimates from the DECREASE trials, given that the
non-DECREASE trials represent the true effect of perioperative
beta-blockade. In contrast to Step 2, which assumes no data manipulation
occurred, Step 3 assumes that the DECREASE trials might in fact contain
manipulated data. The estimates from Step 3 provide an indication of the
extent of potential data manipulation in the DECREASE
studies\textsuperscript{4--6,9}.

@@

In order to estimate the number of manipulated data points, we first
estimated the probability of perioperative mortality (in log odds) in
each trial arm for each trail stratum. As such, we estimate mortality
odds four times: once per condition (beta-blocker or control) per trial
type (DECREASE- and non-DECREASE trials). For all combinations of
condition and trial type, we ran a meta-analysis applying similar
methods used in Step 1, resulting in four meta-analytic absolute
mortality estimates with corresponding effect variances (see Table 3;
one estimate per cell). Throughout the simulations, we used the point
estimates (i.e., fixed effect) to simulate genuine- and manipulated
data, but supplemented this by using the more uncertain distribution
estimates (i.e., random effects) as sensitivity analyses.

\begin{longtable}[]{@{}lll@{}}
\caption{Outcome possibilities within a simulated 2 (beta-blocker v
control) by 2 (dead v alive) clinical trial.}\tabularnewline
\toprule
& Dead & Alive\tabularnewline
Beta-blockers & \(n_{11}\) & \(n_{12}\)\tabularnewline
Control & \(n_{21}\) & \(n_{22}\)\tabularnewline
\bottomrule
\end{longtable}

We applied the inversion method to estimate the number of manipulated
data points in the DECREASE trials\textsuperscript{20}. The inversion
method iteratively hypothesizes that \emph{X} out of \emph{N} data
points were manipulated (i.e., \(X={0, 1, ..., N}\)). For each
combination of \emph{X} and trial, we simulated 10000 datasets. Each
simulated dataset contained \emph{X} manipulated data points and
\emph{N-X} genuine data points. For each simulated dataset (exact
simulation procedure in the next paragraph), we determined the
likelihood of the results with

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (across conditions, without replacement) \emph{X} fictitious
  participants that would be the result of data manipulation.
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell as in Table 3). The meta-analytic point estimate was used or
  a population effect was randomly drawn from the meta-analytic effect
  distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Analyzing DECREASE trials to estimate evidence of data manipulation},
            pdfauthor={Chris H. J. Hartgerink, E. M. Kemper, Markus W. Hollman, Gerben ter Riet},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials to estimate evidence of data manipulation}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{\href{https://orcid.org/}{Chris H. J. Hartgerink}, E. M. Kemper,
\href{https://orcid.org/0000-0001-8248-0244}{Markus W. Hollman},
\href{https://orcid.org/0000-0002-2231-7637}{Gerben ter Riet}}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{27 July, 2017}

@@

\section{Abstract}\label{abstract}

The effect of beta-blockers on perioperative mortality in non-cardiac
surgery has been controversial due to concerns regarding the scientific
integrity of the DECREASE-I and DECREASE-IV trials. In this report, we
investigate these DECREASE trials and model their veracity (i.e., the
probability that these effects or more extreme occurred naturally) and
estimate how many data points might have been manipulated in the
DECREASE trials. Our research indicates that the DECREASE trials are
nearly impossible if we assume they investigate the same effect as the
non-DECREASE trials and under that assumption, our results provide
evidence that at least some data points were manipulated. However, these
differences might also be due to different conceptual approaches as to
how beta-blockade might prevent mortality in non-cardiac surgery.
Considering this, we recommend new and more extensively controlled,
confirmatory trials to determine whether there is any use in
administering beta-blockers in order to decrease perioperative
mortality.

\section{Author note}\label{author-note}

All manuscript materials are available at
\url{https://github.com/chartgerink/2015poldermans} and are preserved at
Zenodo {[}@{]}.

\section{Introduction}\label{introduction}

@@

The trials subject to concerns regarding scientific integrity were the
Dutch DECREASE-I and DECREASE-IV trials\textsuperscript{4--6}. The
committees that investigated the integrity of the DECREASE trials
reported that data manipulation was likely but that the extent of the
data manipulation remained unclear\textsuperscript{4--6}. Moreover, the
latest guidelines still recommend the usage of beta-blockers in the
perioperative period in certain cases\textsuperscript{10,11}, where some
of these guidelines are based on other work by the PI of the DECREASE
trials. Considering the potential harmful consequences of guidelines
based on work by someone who has been repeatedly investigated for
breaching scientific integrity, we aim to estimate the extent of data
manipulation in the DECREASE studies\textsuperscript{2--6} to further
stimulate the debate on using beta-blockers in the perioperative period
for patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data but did not investigate the extent to
which the DECREASE trials deviated from comparable trials. Provenance is
primarily concerned with the origins of the data, verifying things such
as (but not limited to) the informed consent and whether data
corresponded to patient files. The committee reports did not neglect
statistical evaluation however: according to the report a statistical
expert evaluated the applicability of forensic statistical
methods\textsuperscript{6} to evaluate results of trials separately
(i.e., DECREASE-I, DECREASE-IV) although the report lacks details as to
how this evaluation took place. The expert concluded that the previously
applied methods by him were not applicable for use in this case.
Nonetheless, in this report we compare across trials, which is a method
that has previously been used to monitor trial data quality or to test
for potential data anomalies\textsuperscript{12,13}. Moreover, comparing
across trials instead of evaluating them separately has previously
proven to be effective in detecting data
manipulation\textsuperscript{14}. Comparing the DECREASE trials to other
published trials studying the effectiveness of beta-blockers with
respect to perioperative mortality could prove informative of the
potential extent of the manipulation in the DECREASE trials.

@@

To statistically investigate the evidence for data manipulation in the
DECREASE studies\textsuperscript{2,3}, we took three steps. First, we
reproduced the findings from the 2014 meta-analysis by Bouri et
al.\textsuperscript{9}, which contained sufficient information to
estimate the deviation of the DECREASE trials from other published
trials on beta-blockers. We also included type of beta-blocker to
inspect whether this is predictive of the effect of beta-blockers on
perioperative mortality. Second, we evaluated the probability that the
DECREASE trials (or more extreme effects) arose from the same effect
distribution as the non-DECREASE trials, which are assumed to be the
true effect of beta-blockers on perioperative mortality in patients
undergoing non-cardiac surgery. Third, we estimated how many data points
would have to be manipulated in order to reproduce the results of the
DECREASE trials if the initial, non-manipulated results arose from the
effect estimates as obtained from the non-DECREASE trials. Considering
the committees investigating the scientific integrity of the DECREASE
trials was unable to assess this, we consider it worthwhile to
investigate this further.

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled
these using \texttt{v2.0.0} of the \texttt{R} package
\texttt{metafor}\textsuperscript{16}. We estimated a weighted
random-effects model using the restricted maximum-likelihood estimator
(i.e., \texttt{REML})\textsuperscript{17} to estimate the variance of
effects. We used the default weighting procedure in the \texttt{metafor}
package. We added 0.5 to each cell count, as is common in meta-analyses
on risk- and odds ratios in order to prevent computational
artefacts\textsuperscript{18}. The 2014 meta-analysis\textsuperscript{9}
did not specify the variance estimate used, hence, (minor) discrepancies
between our estimates and the original estimates could be due to
differences in the estimation procedure.

@@

We were able to closely reproduce the estimates for the different sets
of studies (Figure 2 of the 2014 meta-analysis\textsuperscript{9}).
Bouri et al. differentiated between the estimates from the non-DECREASE
trials (\(k=9\)) and the DECREASE trials (\(k=2\)). We confirmed the
effect size estimates and the variance estimates for both the
non-DECREASE- and the DECREASE trials, except for some discrepancies at
the second decimal level for the estimated effect sizes and a somewhat
larger difference between the variance estimates of the DECREASE
studies. Table 1 depicts the original and reproduced values for both
sets of studies.

\begin{longtable}[]{@{}lllll@{}}
\caption{The original- and reproduced meta-analytic results based on the
data provided in the 2014 meta-analysis by Bouri et al.}\tabularnewline
\toprule
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
Risk ratio\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
RR Confidence interval\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\(\tau^2\)\strut
\end{minipage}\tabularnewline
Non-DECREASE (k=9) & Original & 1.27 & {[}1.01; 1.60{]} &
0\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
Reproduced\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
1.27\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
{[}1.01; 1.61{]}\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
0\strut
\end{minipage}\tabularnewline
DECREASE (k=2) & Original & 0.42 & {[}0.15; 1.23{]} &
0.29\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
Reproduced\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
0.47\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
{[}0.19; 1.15{]}\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
0.14\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to reproduce results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{9}. Our
results showed a bit more evidence against equal subgroups than the
original meta-analysis\textsuperscript{9} (original:
\(\chi^2(1)=3.91,p=.05\); reproduced: \(\chi^2(1)=6.12,p=0.013\)).
Additionally, the original analyses showed substantial residual
heterogeneity (\(I^2=74.4\)\%) whereas we found no residual
heterogeneity (\(I^2=0\)\%). Different variance estimates (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors (including a reminder after several weeks) but did
not receive a response. Nonetheless, the broad strokes of the
meta-regression confirmed that the DECREASE trials were the determining
predictor for the effectiveness of beta-blockers (including DECREASE:
\(RR=0.509\); excluding DECREASE: \(RR=1.275\)).

Additionally, and exploratively, we evaluated the predictive effect of
the type of beta-blocker used in the trials. Descriptively, the DECREASE
trials remained predictive of decreased mortality (\(RR=0.509\)),
whereas the non-DECREASE trials provide tentative evidence that atenolol
results in lower mortality (\(RR=0.777\)). Nonetheless, for other
beta-blockers in the non-DECREASE trials, there is descriptive evidence
that beta-blockers could increase mortality (bisoprolol: \(RR=2.973\);
metoprolol: \(RR=1.303\); propranolol: \(RR=1.7\)). Table 2 shows the
meta-regression results in full. We do note that the DECREASE studies
only use bisoprolol and any estimates for other beta-blockers are
extrapolations.

\begin{longtable}[]{@{}lll@{}}
\caption{Meta-regression results (k=11) for log(RR), including dummy
predictors for DECREASE trials (reference: non-DECREASE trials) and type
of beta-blockers used in the trial (reference:
atenolol).}\tabularnewline
\toprule
\begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
Estimate\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
95\% CI\strut
\end{minipage}\tabularnewline
Intercept & -0.252 & -1.228; 0.724\tabularnewline
Non-DECREASE & - & -\tabularnewline
DECREASE & -1.765 & -5.05; 1.519\tabularnewline
Atenolol & - & -\tabularnewline
Bisoprolol & 1.342 & -2.015; 4.698\tabularnewline
Metoprolol & 0.517 & -0.49; 1.523\tabularnewline
Propranolol & 0.783 & -1.5; 3.065\tabularnewline
\bottomrule
\end{longtable}

@@

Based on the effect estimates for the non-DECREASE trials from Step 1,
we estimated the probability that the observed effects from the DECREASE
studies (or more extreme) occurred naturally. We assumed that the
non-DECREASE studies estimated the true effect distribution of
perioperative beta-blockade on mortality, not perturbed by publication
bias due to statistical (non)significance. Publication bias was assumed
to not be a problem because a substantial number of nonsignificant
effects are included in the dataset (9 of 11 results are
nonsignificant). Based on this effect distribution, we estimated the
veracity of the DECREASE trials separately, which is the estimated
probability of the observed data under a given true
effect\textsuperscript{19}.

@@

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the two-tailed \(p\)-value
for the null hypothesis that the DECREASE trials arose from the same
effect distribution as the non-DECREASE trials
(\(H_0:\mu_{\bar{x}_1-\bar{x}_2}=0\)). To this end, we applied a Welch
t-test\textsuperscript{20}. As means we used the observed log RR for the
DECREASE trials (i.e., DECREASE-I: -1.44; DECREASE-IV: -0.452) and the
meta-analyzed log RR for the non-DECREASE trials (i.e., 0.243). As
standard deviations we used the standard error for the DECREASE trials
(i.e., DECREASE-I: 0.061; DECREASE-IV: 0.018) and the standard error of
the estimated log RR for the non-DECREASE trials (i.e., 0.002). We
initially preregistered that the DECREASE trials would be regarded as
fixed in the computation of the veracity, which was erroneous because
these also have their own standard error; hence, we applied the Welch
test to take into account the uncertainty in the estimates of both the
DECREASE and non-DECREASE trials.

@@

Figure 1 indicates that the DECREASE trials are highly unlikely under
the estimated effect distribution based on the non-DECREASE trials. More
specifically, the results from DECREASE-I (or more extreme) have a
probability of 1 in 10 000 (\(t(8)=-6.75,p=0.000145\)) and the results
from DECREASE-IV (or more extreme) have a probability of 1 in 1000
(\(t(8)=-4.996,p=0.0010587\)). This indicates that the DECREASE trial
results are unlikely to have come from the same population effect
distribution as the non-DECREASE trials. Moreover, observing two of such
extremely unlikely results jointly, as in the DECREASE trials, would
occur in only 2 out of 10 million sets of two trials (i.e.,
\(0.0000002\)) according to this model. Hence, this result indicates
that the DECREASE trials are severely different from the non-DECREASE
trials.

Results from Step 1 indicated that no between-trial variance (i.e.,
homogeneity; \(\tau^2=0\)) of the effects was observed; given the small
number of trials included (i.e., 9) this estimate is highly uncertain,
however. The total \(N\) across the non-DECREASE trials was 10529. We
conducted sensitivity analyses to see how dependent results are on the
heterogeneity estimate (not preregistered;
\href{https://osf.io/vnmzc}{osf.io/vnmzc}). Fixing the variance estimate
\(\tau^2\) to .5, indicates that the probability of observing the
DECREASE trials jointly is approximately 1.2 out of 100 000 (see Figure
1). To put these numbers into context, a variance of 0.25 would suggest
that results of perioperative beta-blockade vary substantially due to
contextual circumstances of the study, even if perioperative
beta-blockade has no effect whatsoever (RRs between 0.779 and 1.284 in
\textasciitilde{}64\% of the cases).

@@

We estimated the number of data points that would need to be manipulated
to arrive at the estimates from the DECREASE trials, given that the
non-DECREASE trials represent the true effect of perioperative
beta-blockade. In contrast to Step 2, which assumes no data manipulation
occurred and that the DECREASE trials occurred naturally from the same
effect distribution as the non-DECREASE trials, Step 3 assumes that the
DECREASE trials might in fact contain manipulated data. The estimates
from Step 3 provide an indication of the extent of potential data
manipulation in the DECREASE studies\textsuperscript{4--6,9}.

@@

In order to estimate the number of manipulated data points, we first
estimated the probability of perioperative mortality (in log odds) in
each trial arm for each trail stratum. As such, we estimate mortality
odds four times: once per condition (beta-blocker or control) per trial
stratum (DECREASE- and non-DECREASE trials). For all four combinations
of condition and trial type, we ran a meta-analysis applying similar
methods used in Step 1, resulting in four meta-analytic absolute
mortality estimates with corresponding effect variances. Throughout the
simulations, we used the point estimates (i.e., fixed effect) to
simulate genuine- and manipulated data, but supplemented this by using
distribution estimates (i.e., random effects) as sensitivity analyses.

We applied the inversion method to estimate the number of manipulated
data points in the DECREASE trials\textsuperscript{21}. We assumed that
if data are manipulated, each data point is manipulated in the same way
and to the same extent. The inversion method iteratively hypothesizes
that \emph{X} out of \emph{N} data points were manipulated (i.e.,
\(X={0, 1, ..., N}\)), assuming they were manipulated in the same way.
For each combination of \emph{X} and trial, we simulated 10000 datasets.
Each simulated dataset contained \emph{X} manipulated data points and
\emph{N-X} genuine data points. For each simulated dataset (exact
simulation procedure in the next paragraph), we determined the
likelihood of the results with

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (across conditions, without replacement) \emph{X} fictitious
  participants that would be the result of data manipulation.
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell based on the estimates from Table 3). The meta-analytic
  point estimate was used or a population effect was randomly drawn from
  the meta-analytic effect distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}
